---
title: "Q&As"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---

## Introduction

```
You are R software teacher specialising in outbreak analytics and epidemic modelling. Give your answer entirely as R code, including some brief comments explaining key parts of the code. Do not include the original question in your response. Begin your response with a commented line introducing the code.
```

```
NOT USED: Your can use packages from the tidyverse to manage data object.
```

## Q1

Question: How to calculate the weekly growth rate of an outbreak? 

Conditions: 
- Use the first list element from the Simulated Ebola outbreak ebola_sim_clean object from the outbreaks R package. 
- Use the "date of onset" column to calculate the weekly growth rate. 
- Keep the first 20 weeks of the period of the analysis.

### A1

```{r}
# Load required packages
library(outbreaks)
library(i2extras)
library(dplyr)

# Load the simulated Ebola outbreak data
data(ebola_sim_clean)

# Extract the first element of the list
linelist <- ebola_sim_clean[[1]]

# Convert the data to an incidence object
dat <- incidence(
  x = linelist, 
  date_index = "date_of_onset",
  interval = "week"
)

# Filter the dataset to keep the first 20 weeks. 
dat_filter <- dat[1:20,]

# Model the incidence
out <- fit_curve(dat_filter, model = "poisson", alpha = 0.05)

# Print the result
growth_rate(out)

# Plot the result
plot(out)
```

- The `outbreaks` package is loaded to access the simulated Ebola outbreak data.
- The `ebola_sim_clean` object from the package contains the simulated outbreak data.
- The `linelist` object contains the first list element from `ebola_sim_clean`.
- The `incidence()` function converts the data object `linelist` to an `incidence2` class object.
- The `date_of_onset` column is used in the `date_index` argument as the onset dates of the outbreak.
- The `"week"` text string is used to count number of cases per week interval to calculate the weekly growth rate.  
- The `incidence2` object provides observations arranged in descendant order with respect to the `date_index`.
- The number of weeks to consider is set to 20 and stored in the `dat_filter` object.
- The `fit_curve()` function fits a Poisson linear regression model to the epicurve in the `dat_filter` object with an alpha value of `0.05` to calculate the 95% confidence intervals.
- The `growth_rate()` function calculate the weekly growth/decay rate.
- The `plot()` function plot the fitted epicurve.

### Reference

- https://www.reconverse.org/i2extras/articles/fitting_epicurves.html

## Q2

Question: How to generate the effective reproductive number from the growth rate?

1. Is there a relationship between the growth rate and rt?

> i2extras and epitrix

```{r}

```

## Q3

2. How to know if an epidemic is under control? 

> epiestim

need to implement complementary intervention measures?

Can you estimate the time-varying reproduction number (Rt) of the outbreak?

```{r}
library(EpiEstim)

# Import data
data <- read.csv("outbreak_data.csv")

# Format data
data$date <- as.Date(data$date)
cases <- ts(data$cases, start = 1, frequency = 7)

# Estimate R
epiestim <- estimate_R(cases, method="parametric_si",
                   	config=list(mean_si = 5, std_si = 1),
                   	min.cases=10, maxLikelihood = FALSE)

```

## Q4

2. how can I took into account the reporting delays for the rt?

> https://mrc-ide.github.io/EpiEstim/articles/alternative_software.html

```{r}

```

## Q5

1. Can you develop a predictive model to forecast the future trajectory of the outbreak?

> projections
> bpmodels

how is the outbreak likely to develop over the next weeks?
what are the forecasted number of cases over next weeks?
what are the hospital bed requirements for next weeks?

```{r}

```

## Q6

3. superspreading event? compare the population r0 and the individual r0, dispersion parameter

> epicontacts
> bpmodels

> https://community.appliedepi.org/t/estimating-the-degree-of-super-spreading-from-transmission-chain-data/103

```{r}

```

## Q7

2. In which scenarios the epidist object is useful? when to use a report summary statistic vs a dsitribution parameter

> epiparameter

```{r}

```

## Q8

3. How can I create a susceptibility matrix for a combination of for the finalsize package?

> https://epiverse-trace.github.io/finalsize/articles/susceptibility_matrices.html

```{r}

```

## Q9

2. What steps can I follow to create a reproducible report to quantify the time varying transmission during a real-time outbreak evaluation?

> episoap

```{r}

```

