## QX

replace projections with bpmodels

- https://epiverse-trace.github.io/superspreading/articles/superspreading.html
- https://epiverse-trace.github.io/bpmodels/articles/projecting_incidence.html

```{r}
epiparam <- epiparam()

epiparam_ebola_k <- epiparam %>% 
  filter(str_detect(disease,"Ebola")) %>% 
  filter(epi_distribution == "offspring_distribution")

epiparam_ebola_k$mean
```


## Q6

Question: How to generate the effective reproductive number from the growth rate?

1. Is there a relationship between the growth rate and rt?

> i2extras and epitrix

```{r}

```



## Q4

2. how can I took into account the reporting delays for the rt?
  
  > https://mrc-ide.github.io/EpiEstim/articles/alternative_software.html

```{r}

```


## Q6

3. superspreading event? compare the population r0 and the individual r0, dispersion parameter

> epicontacts
> bpmodels

> https://community.appliedepi.org/t/estimating-the-degree-of-super-spreading-from-transmission-chain-data/103

```{r}

```


## Q7

2. In which scenarios the epidist object is useful? when to use a report summary statistic vs a dsitribution parameter

> epiparameter

```{r}

```


## Q9

2. What steps can I follow to create a reproducible report to quantify the time varying transmission during a real-time outbreak evaluation?
  
  > episoap

```{r}

```

## Q2

Question: How to calculate a delay-corrected case fatality ratio?

```{r}

```

## draft

1. Can you get the linelist and contact data from the outbreaks?
2. Can you use the covid regional data from the covidregional package?
3. Pick data before the peak of cases?

How to convert an epidemiological parameter from statistic to distribution?

How to convert an epidemiological parameter from distribution to statistic?

- using a parametrized parametric si
- using a discrete non parametric si --> si_distr
- uncertain si

how is the outbreak likely to develop over the next weeks?
what are the forecasted number of cases over next weeks?
what are the hospital bed requirements for next weeks?