---
title: "How to calculate the final size of an epidemic accounting for heterogeneous social contact?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

## Topic

final size of an epidemic

## Introductory prompt

```{r}
#| child: '../prompts/prompt_01.qmd'
```

```{r}
#| child: '../prompts/prompt_02.qmd'
```

## Question

How to calculate the final size of an epidemic accounting for heterogeneous social contact and homogeneous susceptibility in all age groups?

Conditions:

- (pending)

### Answer

```{r}
#| warning: false
#| eval: false

# Calculate the final size of an epidemic

library(finalsize)
library(socialmixr)
library(tidyverse)

# get UK polymod data
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 5, 18, 40, 65),
  symmetric = TRUE
)

# get the contact matrix and demography data
contact_matrix <- t(contact_data$matrix)

# scale the contact matrix so the largest eigenvalue is 1.0
# this is to ensure that the overall epidemic dynamics correctly reflect
# the assumed value of R0
contact_matrix_scaled <- contact_matrix / max(Re(eigen(contact_matrix)$values))

# Define population in each age group
demography_vector <- contact_data$demography$population
demography_groups <- contact_data$demography$age.group

# divide each row of the contact matrix by the corresponding demography
# this reflects the assumption that each individual in group {j} make contacts
# at random with individuals in group {i}
contact_matrix_random <- contact_matrix_scaled / demography_vector

# Define susceptibility of each group
# susceptibility matrix
susceptibility <- tibble(
  age_group = demography_groups,
  susceptible = c(0.8, 0.8, 0.8, 0.8, 0.8)
) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

susceptibility

# Assume uniform susceptibility within age groups
# demography-in-susceptibility matrix
p_susceptibility <- tibble(
  age_group = demography_groups,
  susceptible = c(1.0, 1.0, 1.0, 1.0, 1.0)
) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

p_susceptibility

# R0 of the disease
r0 <- 1.5 # assumed for pandemic influenza

# Calculate the proportion of individuals infected in each age group
final_size(
  r0 = r0,
  contact_matrix = contact_matrix_random,
  demography_vector = demography_vector,
  susceptibility = susceptibility,
  p_susceptibility = p_susceptibility
)
```

- (pending)

Please note that the code assumes the necessary packages are already installed. If they are not, you can install them using first the `install.packages("pak")` function and then the `pak::pak()` function for both packages in CRAN or GitHub before loading them with `library()`. 

Additionally, make sure to adjust the serial interval distribution parameters according to the specific outbreak you are analyzing.

### Reference

- https://epiverse-trace.github.io/finalsize/articles/susceptibility_matrices.html
- https://epiverse-trace.github.io/finalsize/articles/varying_contacts.html
